---
title: "Robust approaches to modelling fragile conservation biology: flexible solutions to non-normal, hierarchical, temporal, spatial, and phylogenetic data"
author: "Yefeng Yang, Jinming Pan, David Warton, Shinichi Nakagawa"
date: "10-03-2026"
output:
  rmdformats::downcute:
    code_folding: show
    self_contained: true
    thumbnails: false
    lightbox: true
    downcute_theme: "chaos"
pkgdown:
  as_is: true
---


# 0 Code attribution

Parts of the diagnostic workflow and example code in this document are inspired by or adapted from:

Hartig, F. *Advanced Regression Models with R*.  

This work is licensed under a Creative Commons Attribution–NonCommercial–ShareAlike 4.0 International License (CC BY-NC-SA 4.0).

The examples provided here modify and extend the original material by using simulated datasets and workflows tailored to demonstrate (temporal) autocorrelation, heteroscedasticity, and outliers diagnostics.

# 1 Simulate temporally autocorrelated ecological data

We simulate species richness driven by temperature, with AR(1) temporal dependence.

```{r}
set.seed(2026)

AR1sim <- function(n, phi, sigma = 1){
  x <- numeric(n)
  x[1] <- rnorm(1,0,sigma)
  for(i in 2:n){
    x[i] <- phi * x[i-1] + rnorm(1,0,sigma)
  }
  return(x)
}

n <- 1000
year <- 1:n

temperature <- scale(rnorm(n, 15, 3))

#  temporal autocorrelation
epsilon <- AR1sim(n, phi = 0.9, sigma = 0.5)

richness <- 20 + 2*temperature + epsilon

temporal_data <- data.frame(
  year = year,
  temperature = as.numeric(temperature),
  richness = richness
)

plot(temporal_data$year, temporal_data$richness, type="l")
```


# 2 Fit a model ignoring the structured ecological trend

First, let's include any predictors and thus we do not account for any temporal trend. 

```{r}
ignore_trend = lm(richness ~ 1, data = temporal_data)
summary(ignore_trend)
```

Check the residuals:

```{r}
plot(residuals(ignore_trend))
```

Have a look at the correlation between richness as a function of their lag (temporal distance):

```{r}
library(DHARMa)
acf(residuals(ignore_trend), type = "correlation", plot = T)
pacf(residuals(ignore_trend), plot = T)
testTemporalAutocorrelation(ignore_trend, time = temporal_data$year)
```


# 3 Fit a model capturing the ecological trend

This model accounts for temporal trend but does not model temporal dependence.

```{r}
mod_trend = lm(richness ~ temperature, data = temporal_data)
summary(mod_trend)
```


Fit a model accounting for temporal dependence, but not temporal trend:

```{r}
library(nlme)
mod_trend2 <- gls(
  richness ~ temperature,
  data = temporal_data
)
summary(mod_trend2)
```


Given that `gls` is not a `DHARMa` object, we now use `glmmTMB` for illustration:

```{r}
library(glmmTMB)
mod_trend3 = glmmTMB(richness ~ ar1(factor(year) + 0 | 1), data = temporal_data) # `1` can be replaced by `grp` indicating multiple time series
summary(mod_trend3)
testTemporalAutocorrelation(mod_trend3, time = temporal_data$year)
```


Fit a model accounting for both temporal trend and temporal dependence:

```{r}
# using gls
mod_trend4 <- gls(
  richness ~ temperature, 
  corr = corAR1(0.9, form = ~ 1),
  data = temporal_data
)
summary(mod_trend4)

#using glmmTMB
mod_trend5 = glmmTMB(richness ~ temperature + ar1(0 + factor(year) | 1), data = temporal_data)
summary(mod_trend5)
```


# 4 Residual diagnostics with DHARMa

The diagnostic plots evaluate whether residuals behave as expected under the fitted model:

We look for

- uniform residual distribution

- no strong structure in residual plots

- no systematic trends

If temporal autocorrelation exists, residuals often show structured deviations from randomness.

For the diagnostic of the model accounting for temporal trend but does not model temporal dependence (`mod_trend`):

```{r}
acf(residuals(mod_trend), type = "correlation", plot = T)
pacf(residuals(mod_trend), plot = T)
library(DHARMa)
res <- simulateResiduals(mod_trend)
plot(res)
testTemporalAutocorrelation(res, time = temporal_data$year)
```

For the diagnostic of the model accounting for temporal dependence but does not model temporal trend (`mod_trend3`):

```{r}
acf(residuals(mod_trend3), type = "correlation", plot = T)
pacf(residuals(mod_trend3), plot = T)
res <- simulateResiduals(mod_trend3)
plot(res)
testTemporalAutocorrelation(res, time = temporal_data$year)
```

For the diagnostic of the model accounting for both temporal dependence temporal trend (`mod_trend3`):

```{r}
acf(residuals(mod_trend5), type = "correlation", plot = T)
pacf(residuals(mod_trend5), plot = T)
res <- simulateResiduals(mod_trend5)
plot(res)
testTemporalAutocorrelation(res, time = temporal_data$year)
```


**Important interpretation principle:**

We do not recommend relying solely on the P-value obtained from the `DHARMa` package. We encourage evaluate:

- magnitude of the autocorrelation statistic

- residual plots

- ecological plausibility


# 5 Diagnosing heteroscedasticity

Let's visually and statistically diagonise potential heteroscedasticity of the fitted model:

```{r}
# visual
plotResiduals(res, temporal_data$temperature)
# statistical
testDispersion(res)
```

# 6 Detecting outliers

Let's visually and statistically diagonise potential outliers  of the fitted model:

```{r}
# visual
plot(res, quantreg = TRUE)
# statistical
testOutliers(res)
```


# Declaration

This document contains adapted material inspired by examples from
Florian Hartig's *Advanced Regression Models with R*.

Licensed under Creative Commons Attribution–NonCommercial–ShareAlike 4.0.
Adaptations in this document include new simulated datasets (completely different data-generation process), modified
code examples, and integration with DHARMa-based diagnostics.

# Package information

```{r}
subset(data.frame(sessioninfo::package_info()), attached==TRUE, c(package, loadedversion))
```

