---
title: "Robust modelling and inference for dependent data: a practical hands-on tutorial"
author: "Yefeng Yang, Jinming Pan, David Warton, Shinichi Nakagawa"
date: "Aug 2026"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    df_print: paged
    self_contained: true
pkgdown:
  as_is: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  fig.width = 9.2,
  fig.height = 5.8,
  dpi = 120,
  collapse = TRUE,
  comment = "#>"
)

set.seed(2026)

site_highlight <- "#287d8e"
site_grey <- "grey72"
warning_colour <- "#b15f00"
signal_colour <- "#2f6f4e"
model_colour <- "#6b5b95"

has_pkg <- function(pkg) {
  requireNamespace(pkg, quietly = TRUE)
}

format_p <- function(p) {
  ifelse(is.na(p), NA_character_, ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))
}

tidy_coef <- function(model) {
  if (inherits(model, "gls")) {
    tab <- summary(model)$tTable
  } else if (inherits(model, "glmmTMB")) {
    tab <- summary(model)$coefficients$cond
  } else {
    tab <- coef(summary(model))
  }

  data.frame(
    term = rownames(tab),
    estimate = round(tab[, 1], 3),
    se = round(tab[, 2], 3),
    statistic = round(tab[, 3], 2),
    p_value = format_p(tab[, ncol(tab)]),
    row.names = NULL
  )
}

extract_term <- function(model, term, model_name) {
  tab <- tidy_coef(model)
  tab <- tab[tab$term == term, , drop = FALSE]
  if (!nrow(tab)) {
    return(data.frame(
      model = model_name,
      estimate = NA_real_,
      se = NA_real_,
      statistic = NA_real_,
      p_value = NA_character_
    ))
  }
  data.frame(
    model = model_name,
    estimate = tab$estimate,
    se = tab$se,
    statistic = tab$statistic,
    p_value = tab$p_value
  )
}

make_ar1 <- function(n, phi = 0.65, sigma = 1) {
  x <- numeric(n)
  x[1] <- rnorm(1, 0, sigma / sqrt(1 - phi^2))
  for (i in 2:n) {
    x[i] <- phi * x[i - 1] + rnorm(1, 0, sigma)
  }
  x
}

lag_cor_by_group <- function(data, residual, group = "site", time = "year",
                             max_lag = 6, require_regular = TRUE) {
  stopifnot(all(c(residual, group, time) %in% names(data)))
  data <- data[order(data[[group]], data[[time]]), , drop = FALSE]
  pieces <- split(data, data[[group]], drop = TRUE)

  if (require_regular) {
    irregular <- vapply(pieces, function(z) {
      gaps <- diff(z[[time]])
      length(gaps) > 0L && length(unique(gaps)) > 1L
    }, logical(1))
    if (any(irregular)) {
      stop(
        "lag_cor_by_group() uses observation-order lags and requires regular time steps. ",
        "Use a continuous-time or distance-based diagnostic for irregular series.",
        call. = FALSE
      )
    }
  }

  summarise_correlations <- function(values) {
    values <- values[is.finite(values)]
    if (!length(values)) {
      return(c(mean = NA_real_, min = NA_real_, max = NA_real_, n = 0))
    }

    # Fisher-z averaging is less biased than an arithmetic mean of correlations.
    clipped <- pmin(pmax(values, -0.999999), 0.999999)
    c(
      mean = tanh(mean(atanh(clipped))),
      min = min(values),
      max = max(values),
      n = length(values)
    )
  }

  out <- lapply(seq_len(max_lag), function(lag) {
    values <- vapply(pieces, function(z) {
      r <- z[[residual]]
      keep <- is.finite(r)
      r <- r[keep]
      if (length(r) <= lag || stats::sd(r) == 0) {
        return(NA_real_)
      }
      stats::cor(r[(lag + 1):length(r)], r[1:(length(r) - lag)])
    }, numeric(1))

    s <- summarise_correlations(values)
    data.frame(
      lag = lag,
      mean_autocorrelation = unname(s["mean"]),
      min_site = unname(s["min"]),
      max_site = unname(s["max"]),
      n_sites = as.integer(s["n"])
    )
  })

  do.call(rbind, out)
}

plot_lag_correlation <- function(tab, main = "Residual autocorrelation by lag") {
  ggplot2::ggplot(tab, ggplot2::aes(x = lag, y = mean_autocorrelation)) +
    ggplot2::geom_hline(yintercept = 0, linetype = "dashed", colour = "grey45") +
    ggplot2::geom_linerange(
      ggplot2::aes(ymin = min_site, ymax = max_site),
      colour = "grey55",
      linewidth = 0.8
    ) +
    ggplot2::geom_line(colour = site_highlight, linewidth = 0.8) +
    ggplot2::geom_point(size = 2.4, colour = site_highlight) +
    ggplot2::labs(
      title = main,
      x = "Lag in years",
      y = "Mean within-site residual correlation"
    ) +
    theme_tutorial()
}

plot_lag_correlation_base <- function(tab, main = "Residual autocorrelation by lag") {
  graphics::plot(
    tab$lag,
    tab$mean_autocorrelation,
    type = "b",
    pch = 19,
    ylim = range(c(tab$min_site, tab$max_site, 0), na.rm = TRUE),
    xlab = "Lag in years",
    ylab = "Mean within-site residual correlation",
    main = main
  )
  graphics::abline(h = 0, lty = 2, col = "grey45")
  graphics::segments(
    x0 = tab$lag,
    y0 = tab$min_site,
    x1 = tab$lag,
    y1 = tab$max_site,
    col = "grey55"
  )
  invisible(tab)
}

rmse <- function(observed, predicted) {
  sqrt(mean((observed - predicted)^2, na.rm = TRUE))
}

fit_lm_predict <- function(train, test) {
  fit <- lm(
    richness ~ temp_anomaly + habitat_quality + poly(year_c, 2, raw = TRUE),
    data = train
  )
  stats::predict(fit, newdata = test)
}

skip_note <- function(pkg) {
  sprintf(
    "<p class='skip-note'><strong>Not run on this machine:</strong> the <code>%s</code> package is not installed. The code is shown so readers can run it after installing the package.</p>",
    pkg
  )
}

theme_tutorial <- function(base_size = 13) {
  ggplot2::theme_minimal(base_size = base_size) +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      panel.grid.major = ggplot2::element_line(colour = "grey90", linewidth = 0.3),
      plot.title = ggplot2::element_text(face = "bold", colour = "#1f2933"),
      plot.subtitle = ggplot2::element_text(colour = "#4b5563"),
      plot.title.position = "plot",
      axis.title = ggplot2::element_text(colour = "#1f2933"),
      strip.text = ggplot2::element_text(face = "bold", colour = "#1f2933"),
      legend.position = "bottom"
    )
}

needed_pkgs <- function(pkgs) {
  missing <- pkgs[!vapply(pkgs, has_pkg, logical(1))]
  if (length(missing)) {
    stop(
      "Please install the following package(s) before rendering this core section: ",
      paste(missing, collapse = ", "),
      call. = FALSE
    )
  }
}

needed_pkgs(c("ggplot2", "nlme", "glmmTMB", "MASS", "sandwich", "lmtest"))
```

<script>
(function() {
  var key = "robust-expanded-tutorial-theme";

  function readTheme() {
    try {
      return window.localStorage.getItem(key);
    } catch (error) {
      return null;
    }
  }

  function writeTheme(theme) {
    try {
      window.localStorage.setItem(key, theme);
    } catch (error) {
      return null;
    }
  }

  function preferredTheme() {
    var stored = readTheme();
    if (stored === "light" || stored === "dark") {
      return stored;
    }
    return "light";
  }

  function applyTheme(theme) {
    var nextTheme = theme === "dark" ? "dark" : "light";
    document.documentElement.setAttribute("data-theme", nextTheme);
    document.documentElement.style.colorScheme = nextTheme;

    var button = document.getElementById("theme-toggle");
    if (button) {
      var isDark = nextTheme === "dark";
      button.setAttribute("aria-pressed", String(isDark));
      button.setAttribute(
        "aria-label",
        isDark ? "Switch to day mode" : "Switch to night mode"
      );
      button.setAttribute(
        "title",
        isDark ? "Switch to day mode" : "Switch to night mode"
      );
    }
  }

  applyTheme(preferredTheme());

  window.addEventListener("DOMContentLoaded", function() {
    var button = document.getElementById("theme-toggle");
    applyTheme(preferredTheme());

    if (!button) {
      return;
    }

    button.addEventListener("click", function() {
      var current = document.documentElement.getAttribute("data-theme");
      var nextTheme = current === "dark" ? "light" : "dark";
      writeTheme(nextTheme);
      applyTheme(nextTheme);
    });
  });
})();
</script>

<style>
:root {
  --page-bg: #ffffff;
  --text: #2c2f33;
  --muted: #4b5563;
  --heading: #183642;
  --accent: #287d8e;
  --accent-soft: #e6f4f1;
  --accent-border: #c9e4dd;
  --warning: #b15f00;
  --source: #4a6f3a;
  --hero-bg: linear-gradient(135deg, #f4fbfc 0%, #f7f7ff 58%, #fff7ed 100%);
  --hero-border: #d9e7ea;
  --card-bg: rgba(255, 255, 255, 0.78);
  --card-border: #dce7eb;
  --pill-bg: rgba(255, 255, 255, 0.86);
  --workshop-bg: #f4fbfc;
  --warning-bg: #fff7ed;
  --source-bg: #f5faf2;
  --skip-bg: #f6f6f6;
  --access-bg: #f7f7ff;
  --code-bg: #f2f6f8;
  --code-text: #1f2933;
  --inline-code-bg: #f0f2f4;
  --table-border: #e5e7eb;
  --link: #1f6f8b;
  --toc-bg: rgba(255, 255, 255, 0.96);
  --toc-text: #2c2f33;
  --toc-active-bg: #287d8e;
  --toc-active-text: #ffffff;
  --toggle-bg: #ffffff;
  --toggle-track: #e6edf0;
  --toggle-border: #cdd8dd;
  --toggle-thumb: #ffffff;
  --toggle-icon: #6b7280;
  --toggle-icon-active: #183642;
}
html[data-theme="dark"] {
  --page-bg: #111315;
  --text: #c9c1b6;
  --muted: #a9a196;
  --heading: #c9bba5;
  --accent: #e3bf75;
  --accent-soft: #26352f;
  --accent-border: #49615a;
  --warning: #e9ad65;
  --source: #93b47d;
  --hero-bg: linear-gradient(135deg, #14191c 0%, #171820 58%, #211d18 100%);
  --hero-border: #323841;
  --card-bg: rgba(255, 255, 255, 0.045);
  --card-border: #343b43;
  --pill-bg: rgba(255, 255, 255, 0.07);
  --workshop-bg: #162226;
  --warning-bg: #251d14;
  --source-bg: #182318;
  --skip-bg: #1c1f22;
  --access-bg: #1b1b28;
  --code-bg: #1b2228;
  --code-text: #ddd7ce;
  --inline-code-bg: #23272c;
  --table-border: #333941;
  --link: #e3bf75;
  --toc-bg: rgba(17, 19, 21, 0.96);
  --toc-text: #c9c1b6;
  --toc-active-bg: #40392c;
  --toc-active-text: #f2dfb6;
  --toggle-bg: #16191c;
  --toggle-track: #252a30;
  --toggle-border: #464d55;
  --toggle-thumb: #e3bf75;
  --toggle-icon: #8f8a82;
  --toggle-icon-active: #f2dfb6;
}
body {
  font-size: 18px;
  line-height: 1.64;
  background: var(--page-bg);
  color: var(--text);
}
html[data-theme="light"] body,
html[data-theme="dark"] body {
  background: var(--page-bg);
  color: var(--text);
}
a {
  color: var(--link);
}
a:hover,
a:focus {
  color: var(--accent);
}
.main-container {
  max-width: 1260px !important;
}
@media (min-width: 1200px) {
  .container {
    width: 1260px;
  }
}
h1, h2, h3 {
  color: var(--heading);
  line-height: 1.18;
}
.tutorial-hero {
  background: var(--hero-bg);
  border: 1px solid var(--hero-border);
  border-radius: 8px;
  padding: 22px 24px;
  margin: 18px 0 24px 0;
}
.tutorial-hero .eyebrow {
  color: var(--accent);
  font-size: 12px;
  font-weight: 700;
  letter-spacing: 0.08em;
  text-transform: uppercase;
  margin-bottom: 5px;
}
.tutorial-hero .hero-title {
  font-size: 24px;
  font-weight: 700;
  margin: 2px 0 8px 0;
  color: var(--heading);
}
.tutorial-hero .hero-copy {
  max-width: 920px;
  margin-bottom: 0.9rem;
}
.hero-meta {
  display: flex;
  flex-wrap: wrap;
  gap: 8px;
  margin: 16px 0 4px 0;
}
.hero-meta span {
  background: var(--pill-bg);
  border: 1px solid var(--card-border);
  border-radius: 999px;
  color: var(--heading);
  font-size: 13px;
  font-weight: 600;
  padding: 4px 9px;
}
.workflow-grid {
  display: grid;
  grid-template-columns: repeat(auto-fit, minmax(190px, 1fr));
  gap: 12px;
  margin-top: 16px;
}
.workflow-step {
  background: var(--card-bg);
  border: 1px solid var(--card-border);
  border-radius: 8px;
  padding: 16px 17px;
}
.workflow-step strong {
  display: block;
  color: var(--heading);
  margin-bottom: 6px;
}
.workflow-step span {
  color: var(--muted);
  font-size: 15px;
}
.path-label {
  display: inline-block;
  border-radius: 999px;
  padding: 3px 9px;
  font-size: 14px;
  font-weight: 700;
  color: var(--heading);
  background: var(--accent-soft);
  border: 1px solid var(--accent-border);
  margin-right: 0.25rem;
}
.workshop-box {
  border-left: 5px solid var(--accent);
  background: var(--workshop-bg);
  padding: 18px 20px;
  margin: 22px 0;
  font-size: 18px;
  line-height: 1.62;
}
.warning-box {
  border-left: 5px solid var(--warning);
  background: var(--warning-bg);
  padding: 18px 20px;
  margin: 22px 0;
  font-size: 17px;
}
.source-box {
  border-left: 5px solid var(--source);
  background: var(--source-bg);
  padding: 18px 20px;
  margin: 22px 0;
  font-size: 17px;
}
.skip-note {
  border-left: 5px solid #8b8b8b;
  background: var(--skip-bg);
  padding: 16px 18px;
  font-size: 16px;
}
.accessibility-box {
  border-left: 5px solid #5b5b9a;
  background: var(--access-bg);
  padding: 18px 20px;
  margin: 22px 0;
  font-size: 17px;
}
.small {
  font-size: 16px;
}
table {
  font-size: 16px;
  line-height: 1.5;
}
table,
.table,
.table > thead > tr > th,
.table > tbody > tr > th,
.table > tfoot > tr > th,
.table > thead > tr > td,
.table > tbody > tr > td,
.table > tfoot > tr > td {
  border-color: var(--table-border);
  padding: 9px 11px;
}
caption {
  color: var(--muted);
  font-weight: 600;
  font-size: 16px;
  line-height: 1.45;
  padding: 9px 0;
}
pre {
  border-radius: 8px;
  background: var(--code-bg);
  color: var(--code-text);
  border: 1px solid var(--table-border);
  box-shadow: 0 1px 0 rgba(24, 54, 66, 0.04);
  font-size: 16px;
  line-height: 1.58;
  padding: 16px;
  white-space: pre-wrap;
  word-break: normal;
}
pre code,
code {
  color: var(--code-text);
  background: var(--inline-code-bg);
  font-size: 0.95em;
}
pre code {
  background: transparent;
  font-size: inherit;
  line-height: inherit;
}
blockquote,
details {
  color: var(--text);
}
.btn,
.btn-default,
.dropdown-menu {
  background: var(--toggle-bg);
  border-color: var(--toggle-border);
  color: var(--text);
}
.tocify {
  background: var(--toc-bg);
  border-color: var(--table-border);
  font-size: 16px;
}
.tocify .list-group-item {
  background: transparent;
  color: var(--toc-text);
  border-color: transparent;
  padding: 8px 12px;
}
.tocify .list-group-item.active,
.tocify .list-group-item.active:hover,
.tocify .list-group-item.active:focus {
  background: var(--toc-active-bg);
  color: var(--toc-active-text);
  border-color: var(--toc-active-bg);
}
.theme-toggle-wrap {
  display: flex;
  justify-content: flex-end;
  margin: 3px 0 12px 0;
}
.theme-toggle {
  display: inline-flex;
  align-items: center;
  gap: 0.55rem;
  padding: 0;
  border: 0;
  background: transparent;
  color: var(--muted);
  cursor: pointer;
}
.theme-toggle:focus-visible {
  outline: 2px solid var(--accent);
  outline-offset: 4px;
  border-radius: 999px;
}
.theme-toggle-track {
  position: relative;
  display: inline-flex;
  align-items: center;
  justify-content: space-between;
  width: 60px;
  height: 30px;
  padding: 0 8px;
  border: 1px solid var(--toggle-border);
  border-radius: 999px;
  background: var(--toggle-track);
  box-shadow: inset 0 1px 2px rgba(0, 0, 0, 0.12);
}
.theme-toggle-icon {
  position: relative;
  z-index: 2;
  display: inline-block;
  width: 14px;
  height: 14px;
  color: var(--toggle-icon);
}
.theme-toggle-icon.moon::before {
  content: "";
  position: absolute;
  inset: 1px;
  border-radius: 50%;
  background: currentColor;
}
.theme-toggle-icon.moon::after {
  content: "";
  position: absolute;
  top: -1px;
  left: 5px;
  width: 13px;
  height: 13px;
  border-radius: 50%;
  background: var(--toggle-track);
}
.theme-toggle-icon.sun::before {
  content: "";
  position: absolute;
  inset: 3px;
  border-radius: 50%;
  background: currentColor;
}
.theme-toggle-icon.sun::after {
  content: "";
  position: absolute;
  inset: 0;
  border-radius: 50%;
  border: 1px solid currentColor;
}
.theme-toggle-thumb {
  position: absolute;
  top: 3px;
  left: 4px;
  width: 22px;
  height: 22px;
  border-radius: 50%;
  background: var(--toggle-thumb);
  box-shadow: 0 1px 5px rgba(0, 0, 0, 0.24);
  transition: transform 180ms ease;
}
html[data-theme="dark"] .theme-toggle-thumb {
  transform: translateX(29px);
}
html[data-theme="light"] .theme-toggle-icon.sun,
html[data-theme="dark"] .theme-toggle-icon.moon {
  color: var(--toggle-icon-active);
}
.sr-only {
  position: absolute;
  width: 1px;
  height: 1px;
  padding: 0;
  margin: -1px;
  overflow: hidden;
  clip: rect(0, 0, 0, 0);
  white-space: nowrap;
  border: 0;
}
@media (max-width: 900px) {
  body {
    font-size: 17px;
  }
  .main-container {
    max-width: calc(100vw - 28px) !important;
  }
  pre {
    font-size: 0.95rem;
  }
  table {
    display: block;
    overflow-x: auto;
    white-space: nowrap;
  }
}
</style>

<div class="theme-toggle-wrap">
<button id="theme-toggle" class="theme-toggle" type="button" aria-pressed="false" aria-label="Switch to night mode" title="Switch to night mode">
<span class="theme-toggle-track" aria-hidden="true">
<span class="theme-toggle-icon moon"></span>
<span class="theme-toggle-icon sun"></span>
<span class="theme-toggle-thumb"></span>
</span>
<span class="sr-only">Switch between day and night display modes</span>
</button>
</div>

<div class="tutorial-hero">
<div class="eyebrow">Supplementary code tutorial</div>
<div class="hero-title">A practical workflow for dependent biodiversity data</div>
<div class="hero-copy">
Using a simulated wetland monitoring dataset, this tutorial begins with an
oversimplified model and then examines ecological structure, residual dependence,
site effects, heteroscedasticity, influential observations, estimands, robust
uncertainty and prediction. The examples can be run in sequence, with each model
linked to a specific scientific question.
</div>
<div class="hero-meta" aria-label="Tutorial features">
<span>Core examples run from a fresh R session</span>
<span>Ecological interpretation throughout</span>
<span>Captions and alt text</span>
<span>Optional extensions clearly marked</span>
</div>
<div class="workflow-grid" aria-label="Workshop route">
<div class="workflow-step"><strong>1. Examine the pattern</strong><span>Plot richness, residuals and lag correlation before interpreting a slope.</span></div>
<div class="workflow-step"><strong>2. Add ecological structure</strong><span>Separate systematic ecological signal from residual dependence.</span></div>
<div class="workflow-step"><strong>3. Represent dependence</strong><span>Match random effects, correlation and variance structures to the data.</span></div>
<div class="workflow-step"><strong>4. Assess robustness</strong><span>Use diagnostics, sensitivity analyses and blocked validation.</span></div>
</div>
</div>

# Start with the scientific question, not the software

There is no universally best method for dependent data. The appropriate analysis depends first on the scientific question and only then on the available model or software. Two researchers analysing the same monitoring dataset may reasonably choose different approaches because they seek different quantities, predictions or sources of ecological variation.

Before fitting a model, distinguish four related decisions:

1. **What is the goal of the analysis?**  

   An inferential analysis asks what the data suggest about an ecological relationship or model parameter. A predictive analysis asks how accurately outcomes can be predicted for new observations, sites, species or time periods. An exploratory analysis searches for patterns that can motivate more specific questions.

2. **What quantity should the model estimate?**  

   This target quantity is the *estimand*. It might be an average association across a monitoring network, a change within the same wetland through time, a site-specific or species-specific response, variation among species in their responses, or a prediction for an unsampled location. Different models can answer different questions even when they use the same response and predictors.

3. **Why might the observations be dependent?**  

   Measurements may share information because they come from the same site, population, species, observer or study; were collected close together in time or space; or involve species with shared evolutionary history. The analysis should represent the aspects of this structure that matter for the scientific question.

4. **How confidently can the dependence structure be specified?**  
   When the mechanism is scientifically meaningful and sufficiently supported by the data, it may be represented explicitly through random effects, correlation functions or structured latent processes. When the precise structure is uncertain, robust standard errors, dependence-aware resampling and sensitivity analyses can provide complementary protection. These strategies are not mutually exclusive.

## How scientific aims lead to different analytical choices

Consider annual measurements of species richness from the same wetlands. The appropriate starting point depends on what the researcher wants to learn.

| Scientific aim                                                             | Example question                                                                                                        | Target of the analysis                                                             | Useful starting point                                                                                                                   |
|:---------------------------------------------------------------------------|:------------------------------------------------------------------------------------------------------------------------|:---------------------------------------------------------------------------------- |:----------------------------------------------------------------------------------------------------------------------------------------|
| Estimate an ecological association while representing repeated observations| How is richness associated with temperature after accounting for persistent differences among wetlands?                 | A regression association conditional on the specified site and temporal structure  | LMM or GLMM, potentially combined with a temporal correlation structure; GLS for Gaussian responses with structured residual covariance |
| Estimate an average association across a population                        | What is the average change in richness across the monitoring network for a one-unit increase in temperature?            | A population-averaged association                                                  | GEE                                                                                                                                     |
| Estimate change within the same ecological unit                            | Are years warmer than a wetland’s usual conditions associated with lower richness within that wetland?                  | A within-wetland association                                                       | Fixed-effects model or a within-between model                                                                                           |
| Quantify ecological heterogeneity                                          | How much does the association between temperature and richness vary among wetlands or species?                          | Variation in slopes among ecological units                                         | Random-slope LMM or GLMM                                                                                                                |
| Estimate a dependence parameter                                            | How persistent are unexplained departures from the expected richness trajectory between neighbouring years?             | Temporal correlation after accounting for the systematic component                 | GLS, LMM or GLMM with an autoregressive structure                                                                                       |
| Protect inference when the dependence model is uncertain                   | Is the estimated temperature association still supported without assuming an exact within-wetland covariance structure? | The coefficient from the fitted model with dependence-robust uncertainty           | Cluster-robust variance estimation or dependence-aware resampling                                                                       |
| Predict new observations                                                   | How accurately can richness be forecast for future years or unsampled wetlands?                                         | Out-of-sample predictive performance                                               | A dependence-aware predictive model evaluated using temporally or spatially blocked validation                                          |

These are starting points rather than rigid rules. For example, an LMM can estimate average regression coefficients, variation among wetlands and wetland-specific predictions within the same model. Robust variance estimation changes how uncertainty is calculated but does not by itself change the coefficient being estimated. Biological interpretation depends on the complete model, including its link function, random effects or covariance structure, and the population to which the conclusion applies.

## Write down the target before running the code

Before selecting a package or copying a model formula, complete five statements:

1. **My scientific question is:** ...

2. **The quantity I want to estimate or predict is:** ...

3. **My observations may be dependent because:** ...

4. **I want my conclusions to apply to:** ...

5. **I will judge the analysis by:** ...

The fourth statement defines the intended scope of inference. A conclusion about the wetlands observed is not necessarily a conclusion about all wetlands, and a model that predicts future observations from monitored sites may not predict an unsampled site. The fifth statement distinguishes goals such as calibrated uncertainty, estimation of ecological heterogeneity and out-of-sample prediction.

Each model in the tutorial is therefore introduced by asking:

> **What scientific question does this model answer, and what assumptions are required
> for that answer?**

The aim is to connect the scientific question, estimand, dependence structure, inferential method and validation design, rather than to identify one universally superior model.

# Setup: how to run this tutorial

This tutorial is a code-along page. The core examples use a small set of packages. Advanced examples are shown so that readers can copy and run them after installing the relevant packages.

From the directory containing the file, render the page with:

```{r render-command, eval=FALSE}
rmarkdown::render("Expanded_tutorial.Rmd")
```

If a package is missing, the tutorial does one of two things:

- Core code uses base R, `ggplot2`, `nlme`, `glmmTMB`, `sandwich`, `lmtest`, `MASS`, and `sessioninfo` where available.

- Advanced chunks are displayed but skipped if their package is unavailable.

Minimal install for the core tutorial: `install.packages(c( "rmarkdown", "knitr", "ggplot2", "nlme", "glmmTMB", "sandwich", "lmtest", "MASS", "sessioninfo"))`

Full install for optional sections: `install.packages(c("DHARMa", "geepack", "plm", "brms", "clubSandwich"))`


On an ordinary laptop, the core examples should render in under a minute after packages are installed. The Bayesian section is intentionally not run by default.

<div class="accessibility-box">
<p><strong>How to use this page.</strong> If these models are new to you, begin with the question and guidance box, run the code immediately below it, and compare the output with the accompanying explanation. The advanced sections are labelled optional, and the core workflow does not require every package in the matrix.</p>
<p><strong>Accessibility choices.</strong> Plots have captions and alt text in the rendered HTML. Tables have captions where they carry an interpretation. Colour is used to help scanning, but important contrasts are also described in text, point shape, line position, or table values.</p>
</div>

Small glossary for the code-along:

- **Systematic structure**: the biological part of the model, such as effects of
  temperature, habitat, site, or year.
- **Residual**: what remains after the fitted model has accounted for its
  systematic structure.
- **Dependence**: shared information among observations, such as repeated surveys
  from the same site.
- **Autocorrelation**: dependence that changes with distance in time, space, or
  phylogeny.
- **Estimand**: the quantity the model is trying to estimate, such as a within-site
  effect or a population-average effect.

<details>
<summary><strong>Check package availability on this machine</strong></summary>

This table is useful for troubleshooting, but you do not need to study it before starting the tutorial. The core workflow runs when the core packages are installed; optional packages are skipped or shown without running.

```{r package-availability}
packages <- data.frame(
  package = c(
    "rmarkdown", "knitr", "nlme", "glmmTMB", "sandwich", "lmtest",
    "MASS", "ggplot2", "sessioninfo", "DHARMa", "geepack", "plm", "brms",
    "clubSandwich"
  ),
  role = c(
    "render page", "tables and knitting", "GLS and variance structures",
    "GLMMs and covariance structures", "robust covariance matrices",
    "coefficient tests", "robust regression", "figures", "session information",
    "simulation residual diagnostics", "GEE marginal models",
    "panel fixed-effects models", "Bayesian models",
    "cluster-robust small-sample corrections"
  )
)
packages$installed <- vapply(packages$package, has_pkg, logical(1))
knitr::kable(
  packages,
  caption = "Packages used in the tutorial and whether they are installed on the rendering machine."
)
```

</details>

Minimum learning path:

1. Run setup and build the simulated dataset.
2. Work through Examples 1 to 6.
3. Add Examples 7 to 11 if they match your data problem.
4. Treat Bayesian, real-data, and extension panels as optional next steps.

Core examples in this page are meant to run by default. Optional examples are
explicitly labelled and either skipped when packages are missing or shown with
`eval = FALSE`.

# Before you use your own data

<div class="workshop-box">
<p><strong>Question:</strong> How do I adapt the tutorial to my dataset?</p>
<p>First map your columns to the concepts used here, then check whether the repeated measurements and time steps suit the proposed model. The template below should produce one row per observation, unambiguous site and year identifiers, and no unexpected duplicates.</p>
<p><strong>Keep in mind:</strong> a discrete AR(1) structure assumes ordered observations and is most straightforward with regular time steps. Check whether missing years represent genuine gaps before fitting <code>corAR1(form = ~ year | site)</code>.</p>
</div>

```{r own-data-template, eval=FALSE}
# Replace these names with the names in your dataset.
my_data <- data.frame(
  site = your_site_column,
  year = your_year_column,
  richness = your_response_column,
  temp_anomaly = your_temperature_column,
  habitat_quality = your_habitat_column
)

# Basic structure checks.
stopifnot(!anyNA(my_data$site))
stopifnot(!anyNA(my_data$year))
stopifnot(!anyNA(my_data$richness))

# One row per site-year combination?
duplicated_rows <- duplicated(my_data[c("site", "year")])
sum(duplicated_rows)

# How many repeated observations per site?
table(my_data$site)

# Are time steps regular within each site?
time_gaps <- by(my_data$year, my_data$site, function(z) sort(unique(diff(sort(z)))))
time_gaps

# If most sites have gaps larger than one time unit, consider continuous-time
# correlation structures, explicit missing years, or a different validation design.
```

<div class="source-box">
<p><strong>Source checks.</strong> Package claims in this tutorial were checked against primary or official documentation where possible: <a href="https://cran.r-project.org/package%3DglmmTMB/vignettes/covstruct.html">glmmTMB covariance structures</a>, <a href="https://cran.r-project.org/web/packages/nlme/nlme.pdf">nlme manual</a>, <a href="https://sandwich.r-forge.r-project.org/reference/vcovCL.html">sandwich vcovCL</a>, <a href="https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html">DHARMa vignette</a>, <a href="https://paulbuerkner.com/brms/reference/ar.html">brms AR terms</a>, and the package pages linked in the capability matrix below.</p>
</div>

# The problem: my residuals are autocorrelated

<div class="workshop-box">
<p><strong>Question:</strong> Can observations collected across times, places, sites, species or studies be treated as independent?</p>
<p>Such observations often share ecological history. Nearby years or sites, repeated surveys of the same wetland, and related species may therefore be more alike than distant observations. The simulated dataset below makes this structure visible before we separate systematic ecological signal from residual dependence.</p>
<p><strong>Keep in mind:</strong> autocorrelation in the raw response does not by itself justify an AR(1) term. The pattern may instead reflect an omitted ecological predictor or trend.</p>
</div>

The worked examples follow a central principle of the Review: dependence should be
handled in relation to the scientific aim, whether inference, prediction or
exploration. We fit the same ecological system in several ways and ask what question
each model answers.

# Signal, distribution, and dependence

<div class="workshop-box">
<p><strong>Question:</strong> What are systematic structure, residuals and residual dependence?</p>
<p>A model targets a quantity such as a mean, median, other location parameter or transformed mean. The four panels below distinguish the observed data, fitted systematic component, residuals and correlation among those residuals. This is a conceptual decomposition rather than a universal equation.</p>
<p><strong>Keep in mind:</strong> in a GLM the systematic component usually acts through a link function, whereas robust or quantile models may target a different location parameter. Normality is not required for the Gauss–Markov optimality result, although it underpins exact small-sample inference in the classical linear model.</p>
</div>

```{r concept-figure, fig.cap="A small teaching example showing observed data, the fitted systematic component, residuals, and residual autocorrelation.", fig.alt="Four-panel plot. Panel 1 shows observed richness through 20 years. Panel 2 overlays a curved fitted systematic component on the same data. Panel 3 shows residuals around zero through time. Panel 4 shows an autocorrelation function for the residuals."}
set.seed(13)
tiny_year <- 1:20
tiny_signal <- 6 + 0.25 * tiny_year - 0.012 * tiny_year^2
tiny_resid <- make_ar1(20, phi = 0.75, sigma = 0.35)
tiny_y <- tiny_signal + tiny_resid
tiny_fit <- lm(tiny_y ~ poly(tiny_year, 2, raw = TRUE))
tiny_fitted <- fitted(tiny_fit)
tiny_model_resid <- residuals(tiny_fit)

old_par <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot(tiny_year, tiny_y, type = "b", pch = 19,
     xlab = "Year", ylab = "Observed richness", main = "1. Observed data")
plot(tiny_year, tiny_y, pch = 19, xlab = "Year", ylab = "Richness",
     main = "2. Systematic component")
lines(tiny_year, tiny_fitted, col = "#287d8e", lwd = 2)
plot(tiny_year, tiny_model_resid, type = "b", pch = 19,
     xlab = "Year", ylab = "Residual", main = "3. Residuals")
abline(h = 0, lty = 2)
acf(tiny_model_resid, main = "4. Residual dependence")
par(old_par)
```

This view is more general than writing the response simply as "mean plus error" and helps separate model specification from the remaining dependence.

# Build the teaching dataset

<div class="workshop-box">
<p><strong>Worked example:</strong> annual species richness is simulated for 12 wetland sites. Temperature and habitat quality affect richness, sites differ in baseline richness, and unmeasured ecological conditions persist through time.</p>
<p>The code creates <code>wetland_monitoring</code>, displays its first rows and plots the site trajectories. Because the data-generating process is known, we can see what is omitted as the analysis develops from a simple model to richer alternatives.</p>
<p><strong>Keep in mind:</strong> simulation provides a controlled teaching example, not evidence that any method performs well under all conditions.</p>
</div>

```{r simulate-wetland-data}
set.seed(2026)

sites <- paste0("W", sprintf("%02d", 1:12))
years <- 2001:2020

wetland_monitoring <- expand.grid(site = sites, year = years)
wetland_monitoring <- wetland_monitoring[
  order(wetland_monitoring$site, wetland_monitoring$year),
]
wetland_monitoring$site <- factor(wetland_monitoring$site)
wetland_monitoring$year_c <- wetland_monitoring$year - mean(years)

annual_temp <- as.numeric(scale(
  seq(-1, 1, length.out = length(years)) +
    make_ar1(length(years), phi = 0.35, sigma = 0.25)
))

site_quality <- rnorm(length(sites), 0, 0.6)
site_effect <- rnorm(length(sites), 0, 2.2)
names(site_quality) <- sites
names(site_effect) <- sites

wetland_monitoring$temp_anomaly <-
  annual_temp[match(wetland_monitoring$year, years)] +
  rnorm(nrow(wetland_monitoring), 0, 0.25)

wetland_monitoring$habitat_quality <-
  site_quality[as.character(wetland_monitoring$site)] -
  0.025 * wetland_monitoring$year_c +
  rnorm(nrow(wetland_monitoring), 0, 0.25)

wetland_monitoring$site_effect <-
  site_effect[as.character(wetland_monitoring$site)]

wetland_monitoring$epsilon_ar1 <- unlist(lapply(
  sites,
  function(z) make_ar1(length(years), phi = 0.70, sigma = 1.1)
))

wetland_monitoring$richness <-
  24 +
  1.7 * wetland_monitoring$temp_anomaly +
  3.8 * wetland_monitoring$habitat_quality -
  0.025 * wetland_monitoring$year_c^2 +
  wetland_monitoring$site_effect +
  wetland_monitoring$epsilon_ar1 +
  rnorm(nrow(wetland_monitoring), 0, 0.35)

wetland_monitoring$richness <- round(pmax(wetland_monitoring$richness, 1), 1)
wetland_monitoring$year_f <- factor(wetland_monitoring$year, levels = years)

knitr::kable(
  head(wetland_monitoring, 10),
  caption = "First ten rows of the simulated wetland monitoring dataset."
)
```

```{r plot-wetland-data, fig.cap="Simulated richness trajectories for 12 wetland sites from 2001 to 2020. The grey lines show all sites, and site W01 is labelled and highlighted only as a reading guide.", fig.alt="Line plot with year on the x-axis and species richness on the y-axis. Grey trajectories show 12 wetland sites varying in baseline richness and shared temporal structure. Site W01 is highlighted with points and a direct label at the end of the line."}
site_cols <- grDevices::hcl.colors(length(sites), palette = "Dark 3")
names(site_cols) <- sites

ggplot2::ggplot(wetland_monitoring, ggplot2::aes(year, richness, group = site)) +
  ggplot2::geom_line(colour = site_grey, linewidth = 0.5) +
  ggplot2::geom_line(
    data = subset(wetland_monitoring, site == "W01"),
    colour = site_highlight,
    linewidth = 0.9
  ) +
  ggplot2::geom_point(
    data = subset(wetland_monitoring, site == "W01"),
    colour = site_highlight,
    size = 1.7
  ) +
  ggplot2::annotate(
    "text",
    x = max(wetland_monitoring$year) + 0.35,
    y = subset(wetland_monitoring, site == "W01")$richness[
      nrow(subset(wetland_monitoring, site == "W01"))
    ],
    label = "W01",
    hjust = 0,
    colour = site_highlight
  ) +
  ggplot2::scale_x_continuous(expand = ggplot2::expansion(mult = c(0.02, 0.12))) +
  ggplot2::labs(
    title = "Wetland richness through time",
    x = "Year",
    y = "Species richness"
  ) +
  theme_tutorial()
```

```{r plot-single-site, fig.cap="One wetland site shown by itself so the year-to-year pattern is easy to see.", fig.alt="Line and point plot for wetland site W01. Species richness changes through time, with neighbouring years often having similar values."}
one_site <- subset(wetland_monitoring, site == "W01")
ggplot2::ggplot(one_site, ggplot2::aes(year, richness)) +
  ggplot2::geom_line(colour = site_highlight, linewidth = 0.9) +
  ggplot2::geom_point(colour = site_highlight, size = 2) +
  ggplot2::labs(
    title = "One wetland site",
    x = "Year",
    y = "Species richness"
  ) +
  theme_tutorial()
```

# Example 1: fit a naive model

<div class="workshop-box">
<p><strong>Question:</strong> What happens if a simple temperature–richness model treats every row as independent?</p>
<p>The ordinary linear model provides a slope, but it omits habitat, persistent site differences and temporal memory. These omissions can alter the estimated association as well as make its uncertainty too optimistic. Inspect both the coefficient table and the within-site residual autocorrelation.</p>
<p><strong>Next:</strong> add plausible ecological structure before deciding that the remaining pattern is a residual-correlation problem. A small p-value is not persuasive when the model or its standard error is poorly specified.</p>
</div>

```{r naive-model}
m_naive <- lm(richness ~ temp_anomaly, data = wetland_monitoring)
knitr::kable(
  tidy_coef(m_naive),
  caption = "Naive linear model that ignores habitat, site differences, and temporal dependence."
)
```

```{r naive-residual-plots, fig.cap="Residual diagnostics for the naive model.", fig.alt="Two-panel figure. The first panel plots residuals through year for all sites, showing structured bands rather than random scatter. The second panel shows the autocorrelation function for site W01."}
wetland_monitoring$resid_naive <- residuals(m_naive)

old_par <- par(mfrow = c(1, 2))
plot(
  wetland_monitoring$year,
  wetland_monitoring$resid_naive,
  col = site_cols[wetland_monitoring$site],
  pch = 19,
  xlab = "Year",
  ylab = "Residual",
  main = "Naive residuals through time"
)
abline(h = 0, lty = 2)
acf(
  subset(wetland_monitoring, site == "W01")$resid_naive,
  main = "ACF for site W01"
)
par(old_par)
```

```{r naive-lag-correlation, fig.cap="Mean within-site residual autocorrelation for the naive model.", fig.alt="Line plot of residual autocorrelation against lag in years. Mean autocorrelation is positive at short lags, with vertical ranges showing variation among sites."}
naive_lag <- lag_cor_by_group(wetland_monitoring, "resid_naive")
knitr::kable(
  round(naive_lag, 3),
  caption = "Within-site residual autocorrelation by lag for the naive model."
)
plot_lag_correlation(naive_lag, "Naive model residual autocorrelation")
```

# Example 2: add the systematic structure first

<div class="workshop-box">
<p><strong>Question:</strong> Does the apparent dependence reflect residual autocorrelation or omitted ecological structure?</p>
<p>Add habitat quality and a quadratic year trend, then compare the temperature coefficient and residual lag correlations with the naive model. Temporal structure in the raw data may represent ecological signal; accounting for that signal can reduce, alter or reveal the dependence left in the residuals.</p>
<p><strong>Next:</strong> model residual autocorrelation directly if a meaningful pattern remains. An AR(1) term should not be used to absorb an omitted predictor or an obvious trend.</p>
</div>

```{r systematic-model}
m_systematic <- lm(
  richness ~ temp_anomaly + habitat_quality + poly(year_c, 2, raw = TRUE),
  data = wetland_monitoring
)
knitr::kable(
  tidy_coef(m_systematic),
  caption = "Linear model after adding habitat quality and a quadratic year trend."
)
```

```{r compare-naive-systematic}
temp_compare <- rbind(
  extract_term(m_naive, "temp_anomaly", "Naive LM"),
  extract_term(m_systematic, "temp_anomaly", "Systematic LM")
)
knitr::kable(
  temp_compare,
  caption = "Temperature-effect estimate before and after adding systematic ecological structure."
)
```

```{r systematic-lag, fig.cap="Residual autocorrelation before and after adding systematic ecological structure.", fig.alt="Two-panel lag-correlation plot. Both models show positive short-lag residual autocorrelation. Adding habitat and year structure changes the residual pattern, but short-lag autocorrelation remains strong enough to model directly."}
wetland_monitoring$resid_systematic <- residuals(m_systematic)
systematic_lag <- lag_cor_by_group(wetland_monitoring, "resid_systematic")

lag_compare <- rbind(
  data.frame(model = "Naive residuals", naive_lag),
  data.frame(model = "After systematic structure", systematic_lag)
)

ggplot2::ggplot(
  lag_compare,
  ggplot2::aes(x = lag, y = mean_autocorrelation, colour = model)
) +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed", colour = "grey45") +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::geom_point(size = 2.2) +
  ggplot2::scale_colour_manual(
    values = c("Naive residuals" = warning_colour,
               "After systematic structure" = site_highlight)
  ) +
  ggplot2::labs(
    title = "Residual autocorrelation remains after adding biology",
    x = "Lag in years",
    y = "Mean within-site residual correlation",
    colour = NULL
  ) +
  theme_tutorial()

knitr::kable(
  round(systematic_lag, 3),
  caption = "Within-site residual autocorrelation by lag after adding systematic structure."
)
```

In this simulated dataset, adding habitat quality and a year trend changes the temperature estimate and clarifies the ecological signal, but it does not remove residual temporal dependence. The lag-1 residual autocorrelation is still about
`r round(systematic_lag$mean_autocorrelation[systematic_lag$lag == 1], 2)`, so the next step is to model residual temporal dependence directly rather than declare the problem solved.

# Example 3: model residual temporal dependence with GLS

<div class="workshop-box">
<p><strong>Question:</strong> How can we represent residual similarity between neighbouring years?</p>
<p>Generalized least squares (GLS) retains the same systematic component while specifying AR(1) residual correlation within each wetland. Compare the coefficient standard errors and inspect the estimated correlation parameter.</p>
<p><strong>Keep in mind:</strong> the AR(1) structure should be biologically plausible and supported by the sampling design. Calling <code>gls()</code> without a <code>correlation =</code> argument does not model temporal dependence.</p>
</div>

```{r gls-ar1}
gls_formula <- richness ~ temp_anomaly + habitat_quality +
  poly(year_c, 2, raw = TRUE)

# Use maximum likelihood when comparing alternative dependence structures.
m_gls_ind_ml <- nlme::gls(
  gls_formula,
  data = wetland_monitoring,
  method = "ML"
)

m_gls_ar1_ml <- nlme::gls(
  gls_formula,
  correlation = nlme::corAR1(form = ~ year | site),
  data = wetland_monitoring,
  method = "ML"
)

knitr::kable(
  AIC(m_gls_ind_ml, m_gls_ar1_ml),
  caption = "Information-criterion comparison of independence and AR(1) residual structures with the same systematic component."
)

# After selecting the dependence structure, refit with REML for estimation.
m_gls_ar1 <- update(m_gls_ar1_ml, method = "REML")

knitr::kable(
  tidy_coef(m_gls_ar1),
  caption = "REML-fitted GLS model with the same systematic terms and AR(1) residual correlation within site."
)
```

```{r gls-phi}
phi_hat <- coef(m_gls_ar1$modelStruct$corStruct, unconstrained = FALSE)
knitr::kable(
  data.frame(parameter = names(phi_hat), estimate = round(as.numeric(phi_hat), 3)),
  caption = "Estimated AR(1) residual correlation parameter from the GLS model."
)
```

```{r compare-lm-gls-se}
se_compare <- rbind(
  extract_term(m_systematic, "temp_anomaly", "Systematic LM"),
  extract_term(m_gls_ar1, "temp_anomaly", "GLS with AR(1)")
)
knitr::kable(
  se_compare,
  caption = "Temperature-effect comparison between the systematic linear model and GLS AR(1) model."
)
```

# Example 4: model site-level dependence

<div class="workshop-box">
<p><strong>Question:</strong> How should repeated measurements from the same wetland be represented?</p>
<p>A random intercept allows each wetland to have its own baseline richness. Fit the Gaussian mixed model and inspect both the fixed effects and site-level variance component.</p>
<p><strong>Keep in mind:</strong> persistent differences among sites are distinct from AR(1) residual memory within a site. A random intercept does not automatically remove temporal autocorrelation.</p>
</div>

```{r site-status, echo=FALSE, results='asis'}
if (!has_pkg("glmmTMB")) cat(skip_note("glmmTMB"))
```

```{r site-model, eval=has_pkg("glmmTMB")}
m_site <- glmmTMB::glmmTMB(
  richness ~ temp_anomaly + habitat_quality + poly(year_c, 2, raw = TRUE) +
    (1 | site),
  data = wetland_monitoring
)

knitr::kable(
  tidy_coef(m_site),
  caption = "Mixed model with a site-level random intercept."
)
summary(m_site)$varcor
```

This connects directly to the random-effect syntax in Table 1 of the Review:
`(1 | site)` means "allow wetland sites to differ in baseline richness."

# Example 5: combine site structure and temporal dependence

<div class="workshop-box">
<p><strong>Question:</strong> Can hierarchical and temporal residual dependence occur together?</p>
<p>Yes. A wetland may have a persistent baseline while also showing year-to-year residual memory. The advanced <code>glmmTMB</code> model below combines a site intercept with an AR(1) time structure; its variance components and correlation parameter describe different sources of dependence.</p>
<p><strong>Keep in mind:</strong> this is a candidate for sensitivity or multiverse analysis, not an automatic final model. Complex covariance structures can overfit small datasets.</p>
</div>

<div class="warning-box">
<p><strong>Advanced model warning.</strong> The GLS model above is the main residual-correlation teaching example. The <code>glmmTMB</code> model below is included because many readers will want to know how site effects and temporal structure can coexist, but it is easier to overfit and harder to diagnose. If the residual standard deviation is near zero, or a variance component looks boundary-like, treat that as a warning to simplify, check sensitivity, and explain the modelling choice.</p>
</div>

The current `glmmTMB` covariance-structure vignette uses syntax of the form
`ar1(time + 0 | group)`. We use a factor year to represent regularly spaced
annual observations within each site.

```{r combined-status, echo=FALSE, results='asis'}
if (!has_pkg("glmmTMB")) cat(skip_note("glmmTMB"))
```

```{r combined-model, eval=has_pkg("glmmTMB")}
m_combined <- glmmTMB::glmmTMB(
  richness ~ temp_anomaly + habitat_quality + poly(year_c, 2, raw = TRUE) +
    (1 | site) + ar1(0 + year_f | site),
  data = wetland_monitoring
)

knitr::kable(
  tidy_coef(m_combined),
  caption = "Mixed model with site-level baseline differences and within-site AR(1) temporal dependence."
)
summary(m_combined)$varcor
```

In `glmmTMB` output, the AR(1) term may appear as a variance component with a
label such as `site.1 year_f2001` and a correlation line marked `(ar1)`. That
label means the model has created a structured time effect within each site; it
does not mean that only the year 2001 was modelled. A very small residual
standard deviation can also appear when much of the remaining within-site
variation has been allocated to the structured AR(1) term. In practice, that is
a reason to simplify the model or examine alternative specifications in a
sensitivity analysis.

If this model fails or is unstable in a reader's dataset, the safer teaching
move is to return to the GLS AR(1) example and then add complexity one piece at
a time.

# Example 6: interpret diagnostics in context

<div class="workshop-box">
<p><strong>Question:</strong> How should residual plots and diagnostic tests inform the analysis?</p>
<p>Examine residuals against fitted values, through time and across lags, asking whether any remaining structure is large enough to affect the scientific conclusion. Diagnostics guide investigation; they are not automatic model-selection rules.</p>
<p><strong>Keep in mind:</strong> statistical significance depends on sample size. Interpret diagnostic tests alongside plots, effect magnitude, model purpose and ecological plausibility rather than adding terms until every test is non-significant.</p>
</div>

```{r diagnostic-plots, fig.cap="Four diagnostic views for the systematic linear model.", fig.alt="Four-panel diagnostic figure. Panels show residuals against fitted values, residuals through time, the autocorrelation function for one site, and mean within-site lag correlation."}
old_par <- par(mfrow = c(2, 2), mar = c(4, 4, 2.5, 1))
plot(fitted(m_systematic), residuals(m_systematic),
     xlab = "Fitted value", ylab = "Residual",
     main = "Residuals versus fitted")
abline(h = 0, lty = 2)
plot(wetland_monitoring$year, residuals(m_systematic),
     col = site_cols[as.character(wetland_monitoring$site)], pch = 19,
     xlab = "Year", ylab = "Residual",
     main = "Residuals through time")
abline(h = 0, lty = 2)
acf(subset(wetland_monitoring, site == "W01")$resid_systematic,
    main = "ACF for site W01")
plot_lag_correlation_base(systematic_lag, "Across-site lag summary")
par(old_par)
```

```{r dharma-status, echo=FALSE, results='asis'}
if (!has_pkg("DHARMa")) cat(skip_note("DHARMa"))
```

```{r dharma-diagnostics, eval=has_pkg("DHARMa") && has_pkg("glmmTMB"), fig.cap="Optional DHARMa simulation residual diagnostics for the combined model.", fig.alt="Diagnostic plots generated by DHARMa for simulated residuals. These plots assess whether residuals behave as expected under the fitted model."}
sim_res <- DHARMa::simulateResiduals(
  fittedModel = m_combined,
  n = 500,
  plot = FALSE
)
plot(sim_res)

# testTemporalAutocorrelation() requires unique time values. Because the dataset
# contains 12 observations per year, aggregate the DHARMa residuals by year for
# a diagnostic of temporal structure shared across sites. Within-site temporal
# dependence is examined separately with the lag summaries above.
sim_res_by_year <- DHARMa::recalculateResiduals(
  sim_res,
  group = wetland_monitoring$year
)
DHARMa::testTemporalAutocorrelation(
  sim_res_by_year,
  time = sort(unique(wetland_monitoring$year))
)
```

DHARMa includes tests for temporal, spatial, and phylogenetic autocorrelation in
simulation residuals, but the diagnostic scale must match the scientific question.
The code above aggregates residuals by calendar year and therefore evaluates
remaining temporal structure shared across sites. It is not a substitute for
within-site checks. Repeated spatial or temporal coordinates likewise require
aggregation or separate diagnostics for the relevant subgroups.

# Example 7: heteroscedasticity

<div class="workshop-box">
<p><strong>Question:</strong> What if residual variance changes with environmental stress?</p>
<p>Changing variance can itself be ecologically informative: stressful conditions may make richness less predictable. The example increases residual spread with temperature anomaly and compares an ordinary model with a GLS variance model.</p>
<p><strong>Keep in mind:</strong> the variance structure must match the predictor scale. <code>varIdent(form = ~ 1 | group)</code> is intended for categorical groups; continuous variance covariates can be represented with structures such as <code>varPower()</code> or <code>varExp()</code>.</p>
</div>

```{r heteroscedastic-data}
set.seed(2027)
dat_het <- wetland_monitoring
dat_het$temp_abs <- abs(dat_het$temp_anomaly) + 0.2

mu_het <-
  24 +
  1.7 * dat_het$temp_anomaly +
  3.8 * dat_het$habitat_quality -
  0.025 * dat_het$year_c^2 +
  dat_het$site_effect

sigma_het <- 0.5 + 1.1 * dat_het$temp_abs
dat_het$richness <- round(pmax(mu_het + rnorm(nrow(dat_het), 0, sigma_het), 1), 1)

het_formula <- richness ~ temp_anomaly + habitat_quality +
  poly(year_c, 2, raw = TRUE) + site

m_het_naive <- lm(het_formula, data = dat_het)
m_het_gls <- nlme::gls(
  het_formula,
  weights = nlme::varPower(form = ~ temp_abs),
  data = dat_het,
  method = "REML"
)

knitr::kable(
  tidy_coef(m_het_naive),
  caption = "Mean model that assumes constant residual variance in the heteroscedastic dataset."
)
knitr::kable(
  tidy_coef(m_het_gls),
  caption = "GLS model with the same systematic component and residual variance changing with temperature anomaly."
)
```

```{r heteroscedastic-plots, fig.cap="Residual spread before and after modelling heteroscedasticity.", fig.alt="Two-panel residual plot. The naive model shows residual spread changing with temperature anomaly. The GLS model uses normalized residuals and shows more even spread."}
het_plot_data <- rbind(
  data.frame(
    temp_anomaly = dat_het$temp_anomaly,
    residual = residuals(m_het_naive),
    model = "Naive LM residual"
  ),
  data.frame(
    temp_anomaly = dat_het$temp_anomaly,
    residual = residuals(m_het_gls, type = "normalized"),
    model = "Normalized GLS residual"
  )
)

ggplot2::ggplot(het_plot_data, ggplot2::aes(temp_anomaly, residual)) +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed", colour = "grey45") +
  ggplot2::geom_point(alpha = 0.7, colour = site_highlight, size = 1.7) +
  ggplot2::facet_wrap(~ model, nrow = 1, scales = "free_y") +
  ggplot2::labs(
    x = "Temperature anomaly",
    y = "Residual",
    title = "Residual spread before and after a variance model"
  ) +
  theme_tutorial()
```

This example provides the practical counterpart to Box 2 of the Review.

# Example 8: outliers and robustness

<div class="workshop-box">
<p><strong>Question:</strong> How should extreme years or sites be handled?</p>
<p>Extreme observations may be data errors, rare ecological events or signs of model misspecification. Add several identifiable extremes, compare ordinary and robust fits, and report whether the estimated temperature association changes materially.</p>
<p><strong>Keep in mind:</strong> sensitivity analysis is usually preferable to automatic deletion. Biologically real extremes should not be removed solely because they exert strong influence on the fitted model.</p>
</div>

```{r outlier-data}
set.seed(2026)
dat_outlier <- wetland_monitoring
outlier_rows <- sample(seq_len(nrow(dat_outlier)), 5)
dat_outlier$richness[outlier_rows] <- dat_outlier$richness[outlier_rows] +
  sample(c(-9, 10, 12), length(outlier_rows), replace = TRUE)
dat_outlier$outlier_added <- seq_len(nrow(dat_outlier)) %in% outlier_rows

m_out_lm <- lm(richness ~ temp_anomaly + habitat_quality, data = dat_outlier)
m_out_robust <- MASS::rlm(richness ~ temp_anomaly + habitat_quality,
                          data = dat_outlier)

outlier_compare <- rbind(
  extract_term(m_out_lm, "temp_anomaly", "Ordinary LM"),
  data.frame(
    model = "Robust regression (MASS::rlm)",
    estimate = round(coef(m_out_robust)["temp_anomaly"], 3),
    se = NA_real_,
    statistic = NA_real_,
    p_value = NA_character_
  )
)
knitr::kable(
  outlier_compare,
  caption = "Sensitivity of the temperature-effect estimate to ordinary versus robust regression."
)
```

```{r outlier-plot, fig.cap="Residual plot with added extreme observations marked by triangles.", fig.alt="Scatter plot of residuals against temperature anomaly. Ordinary observations are shown as circles and added extreme observations are shown as triangles, so the extreme rows are identifiable without relying on colour alone."}
dat_outlier$resid_lm <- residuals(m_out_lm)
dat_outlier$outlier_label <- ifelse(dat_outlier$outlier_added,
                                    "added extreme row", "ordinary row")

ggplot2::ggplot(dat_outlier, ggplot2::aes(temp_anomaly, resid_lm)) +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed", colour = "grey45") +
  ggplot2::geom_point(
    ggplot2::aes(shape = outlier_label, colour = outlier_label),
    size = 2,
    alpha = 0.85
  ) +
  ggplot2::scale_shape_manual(values = c("added extreme row" = 17,
                                         "ordinary row" = 19)) +
  ggplot2::scale_colour_manual(values = c("added extreme row" = warning_colour,
                                          "ordinary row" = site_highlight)) +
  ggplot2::labs(
    title = "Extreme observations are visible",
    x = "Temperature anomaly",
    y = "LM residual",
    shape = NULL,
    colour = NULL
  ) +
  theme_tutorial()
```

The robust fit assesses sensitivity to extreme observations. It does not account
for temporal or clustered dependence, which would require an additional modelling
or inferential strategy.

# Example 9: conditional, marginal, and within-site effects

<div class="workshop-box">
<p><strong>Question:</strong> Do GLMMs, GEEs and fixed-effects models answer the same question?</p>
<p>Not necessarily. Decomposing temperature into within-site and between-site components shows how closely related models can target different estimands. A within-site coefficient describes years when a wetland is warmer than its own average; a between-site coefficient contrasts wetlands with different average temperatures.</p>
<p><strong>Keep in mind:</strong> conditional effects account for the specified site structure, whereas marginal effects average across the population. Differences between estimands are not evidence of a robustness failure unless the models were intended to answer the same question.</p>
</div>

```{r within-site-base}
dat_within <- wetland_monitoring
dat_within$temp_site_mean <- ave(dat_within$temp_anomaly, dat_within$site)
dat_within$temp_within <- dat_within$temp_anomaly - dat_within$temp_site_mean

m_population <- lm(
  richness ~ temp_anomaly + habitat_quality + poly(year_c, 2, raw = TRUE),
  data = dat_within
)

m_within_between <- lm(
  richness ~ temp_within + temp_site_mean + habitat_quality +
    poly(year_c, 2, raw = TRUE),
  data = dat_within
)

estimand_compare <- rbind(
  extract_term(m_population, "temp_anomaly", "Pooled temperature effect"),
  extract_term(m_within_between, "temp_within", "Within-site temperature effect"),
  extract_term(m_within_between, "temp_site_mean", "Between-site temperature contrast")
)

knitr::kable(
  estimand_compare,
  caption = "A within-between decomposition separates different temperature estimands."
)
```

The pooled temperature coefficient mixes two questions: what happens when a site
is warmer than usual, and how warmer sites differ from cooler sites. The
within-between decomposition separates these. This is often called a Mundlak-style
decomposition in panel-data discussions. Because this teaching dataset uses a
Gaussian model with an identity link, conditional and population-average slopes
are often closer in interpretation than they would be in logistic or other
nonlinear-link models. The main contrast here is therefore pooled versus within-
and between-site effects, not a claim that every GLMM and GEE coefficient must
differ.

```{r conditional-model, eval=has_pkg("glmmTMB")}
m_conditional_mundlak <- glmmTMB::glmmTMB(
  richness ~ temp_within + temp_site_mean + habitat_quality +
    poly(year_c, 2, raw = TRUE) + (1 | site),
  data = dat_within
)

cond_terms <- tidy_coef(m_conditional_mundlak)
cond_terms <- cond_terms[
  cond_terms$term %in% c("temp_within", "temp_site_mean", "habitat_quality"),
]

knitr::kable(
  cond_terms,
  caption = "Conditional mixed model with within-site and between-site temperature terms."
)
```

The conditional mixed model asks about temperature after allowing wetland sites
to differ in baseline richness. The within-site term is usually closest to the
question "what happens when this site is warmer than its own average?"

```{r gee-status, echo=FALSE, results='asis'}
if (!has_pkg("geepack")) cat(skip_note("geepack"))
```

```{r gee-model, eval=has_pkg("geepack")}
dat_within <- dat_within[order(dat_within$site, dat_within$year), , drop = FALSE]
stopifnot(!is.unsorted(as.integer(dat_within$site)))

m_gee <- geepack::geeglm(
  richness ~ temp_within + temp_site_mean + habitat_quality +
    poly(year_c, 2, raw = TRUE),
  id = site,
  waves = year,
  corstr = "ar1",
  data = dat_within
)
summary(m_gee)
```

This GEE fit has only 12 independent site clusters. It is included to clarify the
population-averaged estimand and working-correlation syntax, not to support precise
large-sample inference. In an applied analysis, assess whether the number of clusters
is adequate and consider an appropriate small-sample correction where available.

```{r plm-status, echo=FALSE, results='asis'}
if (!has_pkg("plm")) cat(skip_note("plm"))
```

```{r plm-model, eval=has_pkg("plm")}
m_plm <- plm::plm(
  richness ~ temp_anomaly + habitat_quality + poly(year_c, 2, raw = TRUE),
  data = dat_within,
  index = c("site", "year"),
  model = "within"
)
summary(m_plm)
```

# Example 10: robust standard errors

<div class="workshop-box">
<p><strong>Question:</strong> Can inference be made less sensitive to an uncertain within-site correlation structure?</p>
<p>Cluster-robust standard errors adjust uncertainty without changing the fitted ecological model. Compare ordinary and site-clustered standard errors, making sure that the clustering level reflects the sampling process.</p>
<p><strong>Keep in mind:</strong> robust standard errors require enough independent clusters; report the clustering level and use a small-sample correction such as CR2 when available. They do not repair an omitted predictor, an unsuitable response distribution or biased sampling.</p>
</div>

```{r robust-se}
m_lm_cluster <- lm(
  richness ~ temp_anomaly + habitat_quality,
  data = wetland_monitoring
)

ordinary <- tidy_coef(m_lm_cluster)
n_site_clusters <- length(unique(wetland_monitoring$site))
V_cluster <- sandwich::vcovCL(
  m_lm_cluster,
  cluster = wetland_monitoring$site,
  type = "HC1"
)
clustered_test <- lmtest::coeftest(m_lm_cluster, vcov. = V_cluster)
clustered <- data.frame(
  term = rownames(clustered_test),
  estimate = round(clustered_test[, 1], 3),
  se = round(clustered_test[, 2], 3),
  row.names = NULL
)

ordinary <- ordinary[, c("term", "estimate", "se")]
ordinary$model <- "ordinary model-based SE"
clustered$model <- "site-clustered HC1 SE"
robust_compare <- rbind(ordinary, clustered)
robust_compare <- robust_compare[
  robust_compare$term %in% c("temp_anomaly", "habitat_quality"),
  c("model", "term", "estimate", "se")
]
knitr::kable(
  robust_compare,
  caption = paste(
    "Ordinary standard errors versus illustrative site-clustered robust standard errors.",
    "There are", n_site_clusters, "site clusters."
  )
)
```

With only `r n_site_clusters` site clusters, the HC1 cluster-robust standard
errors above are a teaching comparison rather than a basis for asymptotic
p-values. In an applied analysis, prefer a small-sample correction such as CR2
with Satterthwaite degrees of freedom when `clubSandwich` supports the fitted
model.

```{r clubsandwich-status, echo=FALSE, results='asis'}
if (!has_pkg("clubSandwich")) {
  cat(skip_note("clubSandwich"))
} else {
  cat("<p class='skip-note'><strong>Preferred small-sample robust-SE route:</strong> the <code>clubSandwich</code> code below is run because the package is installed.</p>")
}
```

```{r clubsandwich-example, eval=has_pkg("clubSandwich")}
clubSandwich::coef_test(
  m_lm_cluster,
  vcov = "CR2",
  cluster = wetland_monitoring$site,
  test = "Satterthwaite"
)
```

# Example 11: prediction and data leakage

<div class="workshop-box">
<p><strong>Question:</strong> Why can random cross-validation be optimistic for dependent biodiversity data?</p>
<p>Training and test rows may share nearby years or ecological units, so a random split can evaluate an easier task than the intended application. Compare random-row, future-year and new-site holdouts; the resulting RMSE values differ because the prediction targets differ.</p>
<p><strong>Keep in mind:</strong> validation should match deployment. Hold out future years for forecasting, sites for transfer to new sites, and species or clades for taxonomic transfer.</p>
</div>

```{r blocked-validation}
set.seed(2026)

random_test <- sample(seq_len(nrow(wetland_monitoring)),
                      size = floor(0.25 * nrow(wetland_monitoring)))
future_test <- wetland_monitoring$year >= 2017
site_test <- wetland_monitoring$site %in% sample(levels(wetland_monitoring$site), 3)

split_results <- list(
  random_rows = list(train = !seq_len(nrow(wetland_monitoring)) %in% random_test,
                     test = seq_len(nrow(wetland_monitoring)) %in% random_test),
  future_years = list(train = !future_test, test = future_test),
  new_sites = list(train = !site_test, test = site_test)
)

cv_results <- lapply(names(split_results), function(nm) {
  idx <- split_results[[nm]]
  train <- wetland_monitoring[idx$train, ]
  test <- wetland_monitoring[idx$test, ]
  pred <- fit_lm_predict(train, test)
  data.frame(
    validation = nm,
    n_train = nrow(train),
    n_test = nrow(test),
    rmse = round(rmse(test$richness, pred), 3)
  )
})

knitr::kable(
  do.call(rbind, cv_results),
  caption = "Prediction error under random-row, future-year, and new-site validation."
)
```

The prediction model contains no site-specific term. Consequently, the new-site split
changes the target population but does not directly quantify leakage through a fitted
site effect. The comparison is therefore an illustration of task-aligned validation;
models containing site, spatial or other structured effects can show a larger contrast
between random and blocked splits.

# Optional Bayesian example

<div class="workshop-box">
<p><strong>Question:</strong> Do Bayesian models remove the need to consider autocorrelation?</p>
<p>No. Bayesian models can incorporate dependence structures and priors, but the relevant dependence must still be specified. The <code>brms</code> example is displayed rather than run because it requires Stan tooling and may take substantially longer than the core examples.</p>
<p><strong>Keep in mind:</strong> inspect posterior intervals, posterior predictive checks and convergence diagnostics. Priors can regularize a complex model, but neither priors nor Bayesian computation substitute for an appropriate sampling design and dependence structure.</p>
</div>

```{r brms-status, echo=FALSE, results='asis'}
cat("<p class='skip-note'><strong>Shown but not run by default:</strong> Bayesian examples can take minutes to hours and require <code>brms</code> plus Stan tooling. The code below is visible for readers who want to run it deliberately.</p>")
if (!has_pkg("brms")) cat(skip_note("brms"))
```

```{r brms-ar1-example, eval=FALSE}
set.seed(2026)

m_bayes_ar1 <- brms::brm(
  brms::bf(
    richness ~ temp_anomaly + habitat_quality + poly(year_c, 2, raw = TRUE) +
      (1 | site) + brms::ar(time = year, gr = site, p = 1)
  ),
  data = wetland_monitoring,
  family = gaussian(),
  prior = c(
    brms::prior(normal(0, 2), class = "b"),
    brms::prior(exponential(1), class = "sd"),
    brms::prior(exponential(1), class = "sigma"),
    brms::prior(normal(0, 0.5), class = "ar")
  ),
  chains = 4,
  iter = 2000,
  cores = 4
)
```

For phylogenetic models in `brms`, the pattern is to pass a known covariance
matrix with `gr(species, cov = A)` and `data2 = list(A = A)`.

```{r brms-phylo-example, eval=FALSE}
m_bayes_phylo <- brms::brm(
  trait ~ predictor + (1 | brms::gr(species, cov = A)),
  data = comparative_data,
  data2 = list(A = phylogenetic_covariance_matrix),
  family = gaussian()
)
```

# Optional real ecological data sidebar

<div class="workshop-box">
<p><strong>Question:</strong> How does the workflow extend to real ecological count data?</p>
<p>The core tutorial uses simulation so that the data-generating process is known. This optional sidebar instead uses <code>glmmTMB::Salamanders</code> to illustrate a negative-binomial count model with site-level random effects and zero inflation.</p>
<p><strong>Keep in mind:</strong> this dataset illustrates count-data modelling rather than the tutorial's core autocorrelation example. Cite the package and data source, and do not assume that its sampling structure matches that of another monitoring study.</p>
</div>

The `Salamanders` documentation describes repeated counts from 23 stream sites
sampled four times. The documentation asks users to cite Price et al. (2016)
and the Dryad data package Price et al. (2015, DOI
`10.5061/dryad.5m8f6`) as well as `glmmTMB` when using the dataset.

```{r salamanders-status, echo=FALSE, results='asis'}
if (!has_pkg("glmmTMB")) cat(skip_note("glmmTMB"))
```

```{r salamanders-example, eval=has_pkg("glmmTMB")}
data("Salamanders", package = "glmmTMB")
salamanders <- glmmTMB::Salamanders
knitr::kable(
  head(salamanders, 8),
  caption = "First rows of the glmmTMB Salamanders dataset."
)

m_salamanders <- glmmTMB::glmmTMB(
  count ~ mined + (1 | site),
  ziformula = ~ mined,
  family = glmmTMB::nbinom2,
  data = salamanders
)

knitr::kable(
  tidy_coef(m_salamanders),
  caption = "Negative-binomial count model for the Salamanders sidebar."
)
```

# Extension panels: when the core tutorial is not enough

The core examples teach the workflow. The next panels help readers choose the
right next tool when their data have stronger structure than this tutorial can
fit in a short web page.

## Spatial GLMMs

Use these when observations are linked by geographic distance or spatial fields.
Packages include `sdmTMB`, `INLA`, `spaMM`, and `glmmTMB`.

```{r spatial-sketch, eval=FALSE}
mesh <- sdmTMB::make_mesh(dat, xy_cols = c("x", "y"), cutoff = 10)
m_spatial <- sdmTMB::sdmTMB(
  count ~ habitat + depth,
  data = dat,
  mesh = mesh,
  family = nbinom2(),
  spatial = "on",
  spatiotemporal = "ar1",
  time = "year"
)
```

Caveat: distance-based spatial models often assume a particular covariance
shape. Check stationarity, anisotropy, boundaries, and the scale of sampling.

## Stream networks

Use these for rivers and streams where flow connectivity matters. Stream sites
can be close in Euclidean distance but disconnected by flow.

```{r stream-network-sketch, eval=FALSE}
m_stream <- SSN2::ssn_lm(
  richness ~ temperature + canopy,
  ssn.object = stream_network_object,
  tailup_type = "exponential",
  additive = "afvArea"
)
```

Caveat: this is not ordinary two-dimensional spatial autocorrelation. Directional
flow and branching structure are part of the model.

## Lattice SAR/CAR models

Use these when dependence is defined by neighbours among areas, polygons, or
network nodes rather than by continuous distance.

```{r lattice-sketch, eval=FALSE}
m_bym2 <- INLA::inla(
  count ~ habitat + f(area_id, model = "bym2", graph = adjacency_graph),
  family = "poisson",
  data = area_data
)
```

Caveat: adjacency models answer an areal-neighbour question. They are not
automatically appropriate for point-referenced observations.

## Phylogenetic dependence

Use these when species share evolutionary history.

```{r phylo-sketch, eval=FALSE}
A <- ape::vcv.phylo(tree, corr = TRUE)
m_phylo <- brms::brm(
  trait ~ environment + (1 | brms::gr(species, cov = A)),
  data = trait_data,
  data2 = list(A = A),
  family = gaussian()
)
```

Caveat: the tree, branch lengths, and evolutionary model are assumptions, not
decoration.

## Community and JSDM models

Use these when many species are modelled together and residual association among
species is part of the question.

```{r jsdm-sketch, eval=FALSE}
m_gllvm <- gllvm::gllvm(
  y = species_matrix,
  X = environmental_data,
  family = "negative.binomial",
  num.lv = 2,
  method = "VA"
)
```

Caveat: latent variables can reveal residual structure, but they need careful
interpretation and diagnostics.

## Big spatial and spatio-temporal data

Use these when the dataset is too large for dense covariance matrices or needs
spatio-temporal prediction.

```{r big-spatial-sketch, eval=FALSE}
m_frk <- FRK::SRE(
  f = z ~ covariate,
  data = spatial_data,
  basis = FRK::auto_basis(manifold = FRK::plane(), data = spatial_data)
)
```

Caveat: reduced-rank or basis-function models trade exact covariance detail for
scalability and prediction.

## Genomic or relationship matrices

Use these when the covariance among levels is known or estimated from pedigree,
genomic, or other relationship matrices.

```{r relationship-sketch, eval=FALSE}
m_sommer <- sommer::mmer(
  trait ~ environment,
  random = ~ sommer::vsr(species, Gu = relationship_matrix),
  data = trait_data
)
```

Caveat: dense covariance problems can become computationally difficult, and the
relationship matrix must be biologically defensible.

# Package capability matrix

These tables are a map, not a ranking. Beginner suitability refers to first use
by an applied ecologist, not to the quality of the package. The information is
split into narrower tables so it is easier to read on small screens and with
screen readers.

Start here before reading the tables:

- Repeated measurements from plots, sites, studies, observers, or species:
  start with `lme4`, `glmmTMB`, or `nlme`.
- Gaussian residual autocorrelation through time within sites: start with
  `nlme::gls()` and `corAR1()`.
- Non-Gaussian counts, presence-absence data, or zero inflation: start with
  `glmmTMB`, then add structure cautiously.
- Population-average effects from clustered data: consider `geepack`, but state
  that the estimand is marginal.
- Uncertain correlation but many independent clusters: consider robust standard
  errors, preferably with small-sample corrections.
- Spatial fields, streams, phylogenies, or multivariate communities: go to the
  relevant extension panel first, then choose software.

```{r package-matrix}
link <- function(text, url) {
  sprintf("<a href='%s'>%s</a>", url, text)
}

pkg_matrix <- data.frame(
  problem = c(
    "Gaussian repeated measures", "GLMMs and covariance structures",
    "Hierarchical LMMs/GLMMs", "Bayesian multilevel dependence",
    "Marginal correlated data", "Within-unit panel effects",
    "Robust covariance", "Cluster-robust small-sample inference",
    "Simulation residual diagnostics", "Smooth temporal/spatial signal",
    "Stream-network dependence", "Spatial/spatio-temporal GLMMs",
    "Large spatial/spatio-temporal prediction", "Latent Gaussian Bayesian models",
    "Community/JSDM ordination", "Survey/spatio-temporal multivariate models",
    "Spatial GLMMs", "Known relationship matrices"
  ),
  package = c(
    "nlme", "glmmTMB", "lme4", "brms", "geepack", "plm", "sandwich",
    "clubSandwich", "DHARMa", "mgcv", "SSN2", "sdmTMB", "FRK", "INLA",
    "gllvm", "VAST/tinyVAST", "spaMM", "sommer"
  ),
  core_or_extension = c(
    "core", "core", "extension", "extension", "extension", "extension",
    "core", "extension", "extension", "extension", "extension", "extension",
    "extension", "extension", "extension", "extension", "extension",
    "extension"
  ),
  estimand = c(
    "model-based conditional/GLS", "conditional", "conditional",
    "posterior conditional or marginal contrasts", "marginal",
    "within-unit", "same coefficient, adjusted SE",
    "same coefficient, adjusted SE", "diagnostic, not estimand",
    "smooth systematic component", "model-based stream-network effect",
    "spatial/spatio-temporal conditional", "prediction and latent field",
    "Bayesian latent Gaussian", "latent-variable community response",
    "spatio-temporal density or multivariate response", "conditional",
    "mixed-model effects with known covariance"
  ),
  temporal = c(
    "yes", "yes", "limited directly", "yes", "working correlation",
    "panel time", "via clusters", "via clusters", "yes diagnostics",
    "smooths and GAMM tools", "not main target", "yes", "yes", "yes",
    "indirect", "yes", "yes", "possible"
  ),
  spatial = c(
    "yes for selected corStruct", "yes covariance structures",
    "via grouping, not geostatistical by default", "GP/CAR/SAR options",
    "via clusters, not spatial field", "panel units", "via clusters",
    "via clusters", "spatial diagnostic", "smooth spatial terms",
    "stream networks", "yes", "yes", "yes", "latent structure",
    "yes", "yes", "possible with covariance"
  ),
  non_gaussian = c(
    "mostly no", "yes", "yes", "yes", "yes", "mostly linear",
    "depends on fitted model", "depends on fitted model", "yes diagnostics",
    "yes", "some support", "yes", "yes in recent FRK", "yes", "yes",
    "yes", "yes", "mostly linear/mixed"
  ),
  beginner_suitability = c(
    "medium", "medium", "high for random effects", "medium-low", "medium",
    "medium", "medium", "medium", "high for diagnostics", "medium",
    "medium-low", "medium", "medium-low", "low", "medium", "low",
    "medium-low", "medium-low"
  ),
  caveat = c(
    "Gaussian focus; formula details matter",
    "covariance syntax must match data spacing and grouping",
    "does not directly fit residual AR(1) the way nlme does",
    "requires priors and MCMC diagnostics",
    "working correlation is not a mechanistic covariance model",
    "linear panel framing; estimand differs from random effects",
    "adjusts SE, not the model",
    "cluster count and correction matter",
    "diagnostics depend on correct simulation setup",
    "smooths can model signal, not necessarily residual covariance",
    "requires stream-network data objects",
    "mesh/SPDE choices matter",
    "basis choice and support matter",
    "powerful but steep learning curve",
    "latent variables need ecological interpretation",
    "specialized for survey and multivariate spatio-temporal use",
    "specialized syntax and assumptions",
    "dense covariance and p > n settings need care"
  ),
  verified_source = c(
    link("nlme manual", "https://cran.r-project.org/web/packages/nlme/nlme.pdf"),
    link("glmmTMB covariance vignette", "https://cran.r-project.org/package%3DglmmTMB/vignettes/covstruct.html"),
    link("lme4 CRAN", "https://cran.r-project.org/package%3Dlme4"),
    link("brms AR docs", "https://paulbuerkner.com/brms/reference/ar.html"),
    link("geepack manual", "https://cran.r-project.org/web/packages/geepack/refman/geepack.html"),
    link("plm CRAN", "https://cran.r-project.org/package%3Dplm"),
    link("sandwich vcovCL", "https://sandwich.r-forge.r-project.org/reference/vcovCL.html"),
    link("clubSandwich vcovCR", "https://jepusto.github.io/clubSandwich/reference/vcovCR.html"),
    link("DHARMa vignette", "https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html"),
    link("mgcv CRAN", "https://cran.r-project.org/package%3Dmgcv"),
    link("SSN2 docs", "https://usepa.github.io/SSN2/"),
    link("sdmTMB docs", "https://sdmtmb.github.io/sdmTMB/"),
    link("FRK JSS", "https://www.jstatsoft.org/article/view/v098i04"),
    link("R-INLA latent models", "https://sites.google.com/a/r-inla.org/www/models/latent-models"),
    link("gllvm docs", "https://jenniniku.github.io/gllvm/"),
    link("tinyVAST CRAN", "https://cran.r-project.org/package%3DtinyVAST"),
    link("spaMM CRAN", "https://cran.r-project.org/package%3DspaMM"),
    link("sommer CRAN", "https://cran.r-project.org/package%3Dsommer")
  )
)

pkg_orientation <- pkg_matrix[, c(
  "problem", "package", "core_or_extension", "estimand", "beginner_suitability"
)]

pkg_capabilities <- pkg_matrix[, c(
  "package", "temporal", "spatial", "non_gaussian", "caveat", "verified_source"
)]

knitr::kable(
  pkg_orientation,
  escape = FALSE,
  caption = "Package orientation: what problem each package is commonly used for and what estimand it tends to target."
)

knitr::kable(
  pkg_capabilities,
  escape = FALSE,
  caption = "Package capabilities and caveats. Use this table after choosing the scientific question."
)
```

# Final reporting checklist

Copy this checklist into the analysis supplement, lab notebook, or review
response.

- Scientific goal: inference, prediction, exploration, or a combination.
- Response and distribution: Gaussian, count, binary, proportion, zero-inflated,
  heavy-tailed, or another family.
- Systematic structure: predictors, trend terms, offsets, interactions, nonlinear
  terms, and why they are biologically plausible.
- Dependence structure: site, observer, year, space, stream network, species,
  phylogeny, study, or crossed clusters.
- Estimand: conditional, marginal, within-site, predictive, or descriptive.
- Diagnostics: residual plots, lag correlations, simulation residuals where
  appropriate, and practical magnitude.
- Sensitivity checks: alternative dependence structures, robust SEs, blocked
  validation, outlier sensitivity, and distributional alternatives.
- Prediction validation: random split only if the deployment task is random;
  otherwise use temporal, spatial, site, species, clade, or study blocks.
- Software and versions: packages, R version, seeds, and optional package
  availability.
- Limitations: what the model cannot distinguish, what dependence remains
  uncertain, and what the next data collection would improve.

# Session info

```{r session-info}
if (has_pkg("sessioninfo")) {
  sessioninfo::session_info()
} else {
  sessionInfo()
}
```

